---
title: "MTH 562 Project 2"
author: "Mark May and Andrew Walther"
date: "3/21/2019"
output: html_document
---

```{r}
#install.packages("binom")
library(mosaic)
library(RColorBrewer)
library(binom)
cols <- brewer.pal(8, 'Set1')
```

## Introduction

We will analyze the Wald, Wilson, and Agresti-Coull confidence intervals based off of their interval lengths and coverage levels. This will be done with random samples from a binomial distribution with a selection of six different sets of parameters.

# Wald Confidence Interval

## $p = 0.5, n = 20$
```{r}
set.seed(562)
p11 <- 0.5
n11 <- 20

wald_lower1 <- c()
wald_upper1 <- c()

for(i in 1:10000){
X_11 <- rbinom(prob=p11, size=n11, n=1)
intervals11 <- binom.confint(x=X_11, n=n11, conf.level=0.95, method='asymptotic')

wald_lower1[i] <- intervals11[1, 5]
wald_upper1[i] <- intervals11[1, 6]
}

coverage_wald1 <- sum(wald_lower1 <= p11 & wald_upper1 >=p11)/10000

length_wald1 <- wald_upper1 - wald_lower1

boxplot(length_wald1, col=cols[1], xlab='Length of 95% Wald Intervals (n=20, p=0.5)')

coverage_wald1
coverage_wald_diff95_1 <- coverage_wald1 - 0.95
coverage_wald_diff95_1
mean(length_wald1)
```

- The Wald confidence interval with $p = 0.5, n = 20$ has a length of $0.4268$ and coverage level of $0.9616$. This exceeds the desired $95%$ confidence level by $0.0116$.

## $p = 0.5, n = 50$
```{r}
set.seed(562)
p12 <- 0.5
n12 <- 50

wald_lower2 <- c()
wald_upper2 <- c()

for(i in 1:10000){
X_12 <- rbinom(prob=p12, size=n12, n=1)
intervals12 <- binom.confint(x=X_12, n=n12, conf.level=0.95, method='asymptotic')

wald_lower2[i] <- intervals12[1, 5]
wald_upper2[i] <- intervals12[1, 6]
}

coverage_wald2 <- sum(wald_lower2 <= p12 & wald_upper2 >=p12)/10000

length_wald2 <- wald_upper2 - wald_lower2
boxplot(length_wald2, col=cols[1], xlab='Length of 95% Wald Intervals (n=50, p=0.5)')

coverage_wald2
coverage_wald_diff95_2 <- coverage_wald2 - 0.95
coverage_wald_diff95_2
mean(length_wald2)
```

- The Wald confidence interval with $p = 0.5, n = 50$ has a length of $0.2743$ and coverage level of $0.9359$. This falls short of the desired $95%$ confidence level by $0.0141$.

## $p = 0.7, n = 20$
```{r}
set.seed(562)
p13 <- 0.7
n13 <- 20

wald_lower3 <- c()
wald_upper3 <- c()

for(i in 1:10000){
X_13 <- rbinom(prob=p13, size=n13, n=1)
intervals13 <- binom.confint(x=X_13, n=n13, conf.level=0.95, method='asymptotic')

wald_lower3[i] <- intervals13[1, 5]
wald_upper3[i] <- intervals13[1, 6]
}

coverage_wald3 <- sum(wald_lower3 <= p13 & wald_upper3 >=p13)/10000

length_wald3 <- wald_upper3 - wald_lower3
boxplot(length_wald3, col=cols[1], xlab='Length of 95% Wald Intervals (n=20, p=0.7)')

coverage_wald3
coverage_wald_diff95_3<- coverage_wald3 - 0.95
coverage_wald_diff95_3
mean(length_wald3)
```

- The Wald confidence interval with $p = 0.7, n = 20$ has a length of $0.3888$ and coverage level of $0.9488$. This falls short the desired $95%$ confidence level by $0.0012$.

## $p = 0.7, n = 50$
```{r}
set.seed(562)
p14 <- 0.7
n14 <- 50

wald_lower4 <- c()
wald_upper4 <- c()

for(i in 1:10000){
X_14 <- rbinom(prob=p14, size=n14, n=1)
intervals14 <- binom.confint(x=X_14, n=n14, conf.level=0.95, method='asymptotic')

wald_lower4[i] <- intervals14[1, 5]
wald_upper4[i] <- intervals14[1, 6]
}

coverage_wald4 <- sum(wald_lower4 <= p14 & wald_upper4 >=p14)/10000

length_wald4 <- wald_upper4 - wald_lower4
boxplot(length_wald4, col=cols[1], xlab='Length of 95% Wald Intervals (n=50, p=0.7)')

coverage_wald4
coverage_wald_diff95_4 <- coverage_wald4 - 0.95
coverage_wald_diff95_4
mean(length_wald4)
```

- The Wald confidence interval with $p = 0.7, n = 50$ has a length of $0.2509$ and coverage level of $0.9364$. This falls short the desired $95%$ confidence level by $0.0136$.

## $p = 0.9, n = 20$
```{r}
set.seed(562)
p15 <- 0.9
n15 <- 20

wald_lower5 <- c()
wald_upper5 <- c()

for(i in 1:10000){
X_15 <- rbinom(prob=p15, size=n15, n=1)
intervals15 <- binom.confint(x=X_15, n=n15, conf.level=0.95, method='asymptotic')

wald_lower5[i] <- intervals15[1, 5]
wald_upper5[i] <- intervals15[1, 6]
}

coverage_wald5 <- sum(wald_lower5 <= p15 & wald_upper5 >=p15)/10000

length_wald5 <- wald_upper5 - wald_lower5
boxplot(length_wald5, col=cols[1], xlab='Length of 95% Wald Intervals (n=20, p=0.9)')

coverage_wald5
coverage_wald_diff95_5 <- coverage_wald5 - 0.95
coverage_wald_diff95_5
mean(length_wald5)
```

- The Wald confidence interval with $p = 0.9, n = 20$ has a length of $0.2337$ and coverage level of $0.8744$. This falls short the desired $95%$ confidence level by $0.0756$.

## $p = 0.9, n = 50$
```{r}
set.seed(562)
p16 <- 0.9
n16 <- 50

wald_lower6 <- c()
wald_upper6 <- c()

for(i in 1:10000){
X_16 <- rbinom(prob=p16, size=n16, n=1)
intervals16 <- binom.confint(x=X_16, n=n16, conf.level=0.95, method='asymptotic')

wald_lower6[i] <- intervals16[1, 5]
wald_upper6[i] <- intervals16[1, 6]
}

coverage_wald6 <- sum(wald_lower6 <= p16 & wald_upper6 >=p16)/10000

length_wald6 <- wald_upper6 - wald_lower6
boxplot(length_wald6, col=cols[1], xlab='Length of 95% Wald Intervals (n=50, p=0.9)')

coverage_wald6
coverage_wald_diff95_6 <- coverage_wald6 - 0.95
coverage_wald_diff95_6
mean(length_wald6)
```

- The Wald confidence interval with $p = 0.9, n = 50$ has a length of $0.1611$ and coverage level of $0.8771$. This falls short the desired $95%$ confidence level by $0.0729$.

# Wilson Confidence Interval

## $p = 0.5, n = 20$
```{r}
set.seed(562)
p21 <- 0.5
n21 <- 20

wilson_lower1 <- c()
wilson_upper1 <- c()

for(i in 1:10000){
X_21 <- rbinom(prob=p21, size=n21, n=1)
intervals21 <- binom.confint(x=X_21, n=n21, conf.level=0.95, method='wilson')

wilson_lower1[i] <- intervals21[1, 5]
wilson_upper1[i] <- intervals21[1, 6]
}

coverage_wilson1 <- sum(wilson_lower1 <= p21 & wald_upper1 >=p21)/10000

length_wilson1 <- wilson_upper1 - wilson_lower1
boxplot(length_wilson1, col=cols[2], xlab='Length of 95% Wilson Intervals (n=20, p=0.5)')

coverage_wilson1
coverage_wilson_diff95_1 <- coverage_wilson1 - 0.95
coverage_wilson_diff95_1
mean(length_wilson1)
```

- The Wilson confidence interval with $p = 0.5, n = 20$ has a length of $0.3926$ and coverage level of $0.9616$. This exceeds the desired $95%$ confidence level by $0.0116$.

## $p = 0.5, n = 50$
```{r}
set.seed(562)
p22 <- 0.5
n22 <- 50

wilson_lower2 <- c()
wilson_upper2 <- c()

for(i in 1:10000){
X_22 <- rbinom(prob=p22, size=n22, n=1)
intervals22 <- binom.confint(x=X_22, n=n22, conf.level=0.95, method='wilson')

wilson_lower2[i] <- intervals22[1, 5]
wilson_upper2[i] <- intervals22[1, 6]
}

coverage_wilson2 <- sum(wilson_lower2 <= p22 & wilson_upper2 >=p22)/10000

length_wilson2 <- wilson_upper2 - wilson_lower2
boxplot(length_wilson2, col=cols[2], xlab='Length of 95% Wilson Intervals (n=50, p=0.5)')

coverage_wilson2
coverage_wilson_diff95_2 <- coverage_wilson2 - 0.95
coverage_wilson_diff95_2
mean(length_wilson2)
```

- The Wilson confidence interval with $p = 0.5, n = 50$ has a length of $0.2646$ and coverage level of $0.9359$. This falls short of the desired $95%$ confidence level by $0.0141$.

## $p = 0.7, n = 20$
```{r}
set.seed(562)
p23 <- 0.7
n23 <- 20

wilson_lower3 <- c()
wilson_upper3 <- c()

for(i in 1:10000){
X_23 <- rbinom(prob=p23, size=n23, n=1)
intervals23 <- binom.confint(x=X_23, n=n23, conf.level=0.95, method='wilson')

wilson_lower3[i] <- intervals23[1, 5]
wilson_upper3[i] <- intervals23[1, 6]
}

coverage_wilson3 <- sum(wilson_lower3 <= p23 & wilson_upper3 >=p23)/10000

length_wilson3 <- wilson_upper3 - wilson_lower3
boxplot(length_wilson3, col=cols[2], xlab='Length of 95% Wilson Intervals (n=20, p=0.7)')

coverage_wilson3
coverage_wilson_diff95_3 <- coverage_wilson3 - 0.95
coverage_wilson_diff95_3
mean(length_wilson3)
```

- The Wilson confidence interval with $p = 0.7, n = 20$ has a length of $0.3643$ and coverage level of $0.9769$. This exceeds the desired $95%$ confidence level by $0.0269$.

## $p = 0.7, n = 50$
```{r}
set.seed(562)
p24 <- 0.7
n24 <- 50

wilson_lower4 <- c()
wilson_upper4 <- c()

for(i in 1:10000){
X_24 <- rbinom(prob=p24, size=n24, n=1)
intervals24 <- binom.confint(x=X_24, n=n24, conf.level=0.95, method='wilson')

wilson_lower4[i] <- intervals24[1, 5]
wilson_upper4[i] <- intervals24[1, 6]
}

coverage_wilson4 <- sum(wilson_lower4 <= p24 & wilson_upper4 >=p24)/10000

length_wilson4 <- wilson_upper4 - wilson_lower4
boxplot(length_wilson4, col=cols[2], xlab='Length of 95% Wilson Intervals (n=50, p=0.7)')

coverage_wilson4
coverage_wilson_diff95_4 <- coverage_wilson4 - 0.95
coverage_wilson_diff95_4
mean(length_wilson4)
```

- The Wilson confidence interval with $p = 0.7, n = 50$ has a length of $0.2438$ and coverage level of $0.9587$. This exceeds the desired $95%$ confidence level by $0.0087$.

## $p = 0.9, n = 20$
```{r}
set.seed(562)
p25 <- 0.9
n25 <- 20

wilson_lower5 <- c()
wilson_upper5 <- c()

for(i in 1:10000){
X_25 <- rbinom(prob=p25, size=n25, n=1)
intervals25 <- binom.confint(x=X_25, n=n25, conf.level=0.95, method='wilson')

wilson_lower5[i] <- intervals25[1, 5]
wilson_upper5[i] <- intervals25[1, 6]
}

coverage_wilson5 <- sum(wilson_lower5 <= p25 & wilson_upper5 >=p25)/10000

length_wilson5 <- wilson_upper5 - wilson_lower5
boxplot(length_wilson5, col=cols[2], xlab='Length of 95% Wilson Intervals (n=20, p=0.9)')

coverage_wilson5
coverage_wilson_diff95_5 <- coverage_wilson5 - 0.95
coverage_wilson_diff95_5
mean(length_wilson5)
```

- The Wilson confidence interval with $p = 0.9, n = 20$ has a length of $0.0053$ and coverage level of $0.9553$. This exceeds the desired $95%$ confidence level by $0.2629$.

## $p = 0.9, n = 50$
```{r}
set.seed(562)
p26 <- 0.9
n26 <- 50

wilson_lower6 <- c()
wilson_upper6 <- c()

for(i in 1:10000){
X_26 <- rbinom(prob=p26, size=n26, n=1)
intervals26 <- binom.confint(x=X_26, n=n26, conf.level=0.95, method='wilson')

wilson_lower6[i] <- intervals26[1, 5]
wilson_upper6[i] <- intervals26[1, 6]
}

coverage_wilson6 <- sum(wilson_lower6 <= p26 & wilson_upper6 >=p26)/10000

length_wilson6 <- wilson_upper6 - wilson_lower6
boxplot(length_wilson6, col=cols[2], xlab='Length of 95% Wilson Intervals (n=50, p=0.9)')

coverage_wilson6
coverage_wilson_diff95_6 <- coverage_wilson6 - 0.95
coverage_wilson_diff95_6
mean(length_wilson6)
```

- The Wilson confidence interval with $p = 0.9, n = 50$ has a length of $0.0213$ and coverage level of $0.9713$. This exceeds the desired $95%$ confidence level by $0.1665$.

# Agresti-Coull Confidence Interval

## $p = 0.5, n = 20$
```{r}
set.seed(562)
p31 <- 0.5
n31 <- 20

agresti_lower1 <- c()
agresti_upper1 <- c()

for(i in 1:10000){
X_31 <- rbinom(prob=p31, size=n31, n=1)
intervals31 <- binom.confint(x=X_31, n=n31, conf.level=0.95, method='agresti-coull')

agresti_lower1[i] <- intervals31[1, 5]
agresti_upper1[i] <- intervals31[1, 6]
}

coverage_agresti1 <- sum(agresti_lower1 <= p31 & agresti_upper1 >=p31)/10000

length_agresti1 <- agresti_upper1 - agresti_lower1
boxplot(length_agresti1, col=cols[3], xlab='Length of 95% Agresti-Coull Intervals (n=20, p=0.5)')

coverage_agresti1
coverage_agresti_diff95_1 <- coverage_agresti1 - 0.95
coverage_agresti_diff95_1
mean(length_agresti1)
```

- The Agresti-Coull confidence interval with $p = 0.5, n = 20$ has a length of $0.3941$ and coverage level of $0.9616$. This exceeds the desired $95%$ confidence level by $0.0116$.

## $p = 0.5, n = 50$
```{r}
set.seed(562)
p32 <- 0.5
n32 <- 50

agresti_lower2 <- c()
agresti_upper2 <- c()

for(i in 1:10000){
X_32 <- rbinom(prob=p32, size=n32, n=1)
intervals32 <- binom.confint(x=X_32, n=n32, conf.level=0.95, method='agresti-coull')

agresti_lower2[i] <- intervals32[1, 5]
agresti_upper2[i] <- intervals32[1, 6]
}

coverage_agresti2 <- sum(agresti_lower2 <= p32 & agresti_upper2 >=p32)/10000

length_agresti2 <- agresti_upper2 - agresti_lower2
boxplot(length_agresti2, col=cols[3], xlab='Length of 95% Agresti-Coull Intervals (n=50, p=0.5)')

coverage_agresti2
coverage_agresti_diff95_2 <- coverage_agresti2 - 0.95
coverage_agresti_diff95_2
mean(length_agresti2)
```

- The Agresti-Coull confidence interval with $p = 0.5, n = 50$ has a length of $0.2648$ and coverage level of $0.9359$. This falls short of the desired $95%$ confidence level by $0.0141$.

## $p = 0.7, n = 20$
```{r}
set.seed(562)
p33 <- 0.7
n33 <- 20

agresti_lower3 <- c()
agresti_upper3 <- c()

for(i in 1:10000){
X_33 <- rbinom(prob=p33, size=n33, n=1)
intervals33 <- binom.confint(x=X_33, n=n33, conf.level=0.95, method='agresti-coull')

agresti_lower3[i] <- intervals33[1, 5]
agresti_upper3[i] <- intervals33[1, 6]
}

coverage_agresti3 <- sum(agresti_lower3 <= p33 & agresti_upper3 >=p33)/10000

length_agresti3 <- agresti_upper3 - agresti_lower3
boxplot(length_agresti3, col=cols[3], xlab='Length of 95% Agresti-Coull Intervals (n=20, p=0.7)')

coverage_agresti3
coverage_agresti_diff95_3 <- coverage_agresti3 - 0.95
coverage_agresti_diff95_3
mean(length_agresti3)
```

- The Agresti-Coull confidence interval with $p = 0.7, n = 20$ has a length of $0.3708$ and coverage level of $0.9769$. This exceeds the desired $95%$ confidence level by $0.0269$.

## $p = 0.7, n = 50$
```{r}
set.seed(562)
p34 <- 0.7
n34 <- 50

agresti_lower4 <- c()
agresti_upper4 <- c()

for(i in 1:10000){
X_34 <- rbinom(prob=p34, size=n34, n=1)
intervals34 <- binom.confint(x=X_34, n=n34, conf.level=0.95, method='agresti-coull')

agresti_lower4[i] <- intervals34[1, 5]
agresti_upper4[i] <- intervals34[1, 6]
}

coverage_agresti4 <- sum(agresti_lower4 <= p34 & agresti_upper4 >=p34)/10000

length_agresti4 <- agresti_upper4 - agresti_lower4
boxplot(length_agresti4, col=cols[3], xlab='Length of 95% Agresti-Coull Intervals (n=50, p=0.7)')

coverage_agresti4
coverage_agresti_diff95_4 <- coverage_agresti4 - 0.95
coverage_agresti_diff95_4
mean(length_agresti4)
```

- The Agresti-Coull confidence interval with $p = 0.7, n = 50$ has a length of $0.2455$ and coverage level of $0.9587$. This exceeds the desired $95%$ confidence level by $0.0087$.

## $p = 0.9, n = 20$
```{r}
set.seed(562)
p35 <- 0.9
n35 <- 20

agresti_lower5 <- c()
agresti_upper5 <- c()

for(i in 1:10000){
X_35 <- rbinom(prob=p35, size=n35, n=1)
intervals35 <- binom.confint(x=X_35, n=n35, conf.level=0.95, method='agresti-coull')

agresti_lower5[i] <- intervals35[1, 5]
agresti_upper5[i] <- intervals35[1, 6]
}

coverage_agresti5 <- sum(agresti_lower5 <= p35 & agresti_upper5 >=p35)/10000

length_agresti5 <- agresti_upper5 - agresti_lower5
boxplot(length_agresti5, col=cols[3], xlab='Length of 95% Agresti-Coull Intervals (n=20, p=0.9)')

coverage_agresti5
coverage_agresti_diff95_5 <- coverage_agresti5 - 0.95
coverage_agresti_diff95_5
mean(length_agresti5)
```

- The Agresti-Coull confidence interval with $p = 0.9, n = 20$ has a length of $0.2912$ and coverage level of $0.9553$. This exceeds the desired $95%$ confidence level by $0.0053$.

## $p = 0.9, n = 50$
```{r}
set.seed(562)
p36 <- 0.9
n36 <- 50

agresti_lower6 <- c()
agresti_upper6 <- c()

for(i in 1:10000){
X_36 <- rbinom(prob=p36, size=n36, n=1)
intervals36 <- binom.confint(x=X_36, n=n36, conf.level=0.95, method='agresti-coull')

agresti_lower6[i] <- intervals36[1, 5]
agresti_upper6[i] <- intervals36[1, 6]
}

coverage_agresti6 <- sum(agresti_lower6 <= p36 & agresti_upper6 >=p36)/10000

length_agresti6 <- agresti_upper6 - agresti_lower6
boxplot(length_agresti6, col=cols[3], xlab='Length of 95% Agresti-Coull Intervals (n=50, p=0.9)')

coverage_agresti6
coverage_agresti_diff95_6 <- coverage_agresti6 - 0.95
coverage_agresti_diff95_6
mean(length_agresti6)
```

- The Agresti-Coull confidence interval with $p = 0.9, n = 50$ has a length of $0.0213$ and coverage level of $0.9713$. This exceeds the desired $95%$ confidence level by $0.1760$.

## Collected Data
Method|$p$|$n$|Coverage|Error|Length
---|---|---|---|---|---
Wald|0.5|20|0.9616|0.0116|0.4268
Wald|0.5|50|0.9359|-0.0141|0.2743
Wald|0.7|20|0.9488|-0.0012|0.3888
Wald|0.7|50|0.9364|-0.0136|0.2509
Wald|0.9|20|0.8744|-0.0756|0.2337
Wald|0.9|50|0.8771|-0.0729|0.1611
Wilson|0.5|20|0.9616|0.0116|0.3926
Wilson|0.5|50|0.9359|-0.0141|0.2646
Wilson|0.7|20|0.9769|0.0269|0.3643
Wilson|0.7|50|0.9587|0.0087|0.2438
Wilson|0.9|20|0.9553|0.0053|0.2629
Wilson|0.9|50|0.9713|0.0213|0.1665
Agresti-Coull|0.5|20|0.9616|0.0116|0.3941
Agresti-Coull|0.5|50|0.9359|-0.0141|0.2648
Agresti-Coull|0.7|20|0.9769|0.0269|0.3706
Agresti-Coull|0.7|50|0.9587|0.0087|0.2455
Agresti-Coull|0.9|20|0.9553|0.0053|0.2912
Agresti-Coull|0.9|50|0.9713|0.0213|0.1760

## Conclusions

### $p = 0.5, n = 20$
For this set of parameters, all three confidence intervals exceeded the desired coverage level of $95%$ by $0.0116$. The Wilson interval has the shortest length at $0.3926$ and the Wald interval has the largest length at $0.4268$. These results indicate that the Wilson confidence interval is the best performer for this particular set of parameters.

### $p = 0.5, n = 50$
For this set of parameters, all three confidence intervals fell short of the desired coverage level of $95%$ by $0.0141$. The Wilson interval has the shortest length at $0.2646$ and the Wald interval has the largest length at $0.2743$. These results indicate that the Wilson confidence interval is the best performer for this set of parameters.

### $p = 0.7, n = 20$
For this set of parameters, the Wald confidence interval fell short of the desired confidence level by $0.0012$. This is the closest interval to the desired confidence level. The Wilson confidence interval has the shortest length at $0.3643$ and the Wald has the largest length at $0.3888$. Since the Wald interval's coverage is the closest to the desired confidence level, it can be considered the best performer for these parameters.

### $p = 0.7, n = 50$
For this set of parameters, the Wilson and Agresti-Coull intervals have coverage levels closest to the desired confidence level. They both fall short by $0.0087$. The Wilson interval has the shortest length at $0.2438$ and the Wald interval's length is the largest at $0.2509$. These results indicate that the Wilson interval is the best performer for these parameters.

### $p = 0.9, n = 20$
For this set of parameters, the Wilson and Agresti-Coull intervals have coverage levels closest to the desired confidence level. They both exceed $95%$ by $0.0053$. The Wald interval has the shortest length at $0.2337$, the length of the Wilson interval is next at $0.2629$, followed by the Agresti-Coull interval at $0.2912$. With a strong balance of accurate coverage and short length, the Wilson interval is the best performing interval for these parameters.

### $p = 0.9, n = 50$
For this set of parameters, the Wilson and Agresti-Coull intervals have coverages levels that both exceed the desired confidence level of $95%$ by $0.0213$. The Wald interval has the shortest length at $0.1611$, followed by the Wilson at $0.1665$, and the Agresti-Coull at $0.1760$. Again, with a balance of accurate coverage and a short interval length, the Wilson interval is the best performer for these parameters.

### Overall
A brief analysis of the three confidence intervals for six sets of parameters led us to conclude that the Wilson interval is the best performer for five of the six sets of parameters and the Wald interval is the best performer for the final set of parameters. 

- The Wald interval had its best result for coverage when $p = 0.7$ and $n = 20$. The shortest length for the Wald occured when $p = 0.9$ and $n = 50$. In general, the Wald interval has coverage values closest to the desired level when $p$ is small and the shortest interval lengths occur when $n$ is large and $p$ is large. Generally, the Wald's performance on coverage was abysmal with $p=0.9$. This is easily understood in the context of the normality assumption that underlies the Wald confidence interval. As $p \rightarrow 0$ or 1, the estimation of $p$ cannot approach normality because $p$ truncates at 0 or 1, preventing the sampling distribution from reaching the needed symmetry. 

- For the Wilson interval, the best coverage result occured with $p = 0.9$ and $n = 20$. The shortest length for the Wilson occured with $p = 0.9$ and $n = 50$. In general, the Wilson interval had better coverage levels when $p$ is large. Similar to the Wald, the shortest interval lengths for the Wilson occur when $n$ is large and $p$ is large. The Wilson's coverage results did not suffer with a high $p$ as much as the Wald and Agresti-Coull did. This is because the Wilson confidence interval does not require assuming that $p = \hat{p}$, so it is more robust when the normality assumption breaks down.

- The best coverage result for the Agresti-Coull interval occured with $p = 0.9$ and $n = 20$. The shortest interval length occured with $p = 0.9$ and $n = 50$. The Agresti-Coull interval had better coverage levels near $p = 0.9$. The shortest interval lengths for the Agresti-Coull occur when $n$ is large and $p$ is large. The Agresti-Coull performed better than the Wald as $p \rightarrow 1$, which is good because that's what it's supposed to do. The act of using $\tilde{p} = \frac{X+2}{n+4}$ as the estimator for $p$ nudges it away from $0$ and $1$, dampening the disastrous effect those extreme values can have on the Wald confidence interval.

In summation, the simulation results indicate that the Wilson confidence interval has the best overall performance combined between coverage levels near the desired value and short interval lengths. This would lead us to conclude that the Wilson interval is "better" than the Wald and Agresti-Coull intervals.
